### 2014-11-07 doc/thesis/evaluation.tex
Splitting the evaluation chapter into two sections: case studies and related work

We performed some case studies on existing Lua libraries
to evaluate the design of our type system.
For each library, we used Typed Lua to either rewrite its modules
or to write statically typed interfaces to its modules through
Typed Lua's description files.
In this chapter we present our evaluation results, discuss
some interesting cases, and compare our type system to
related work.

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|}
\hline
\textbf{Case study} & \textbf{easy} & \textbf{poly} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
\hline
Lua Standard Libraries 5.2 & 94 & 5 & 34 & 133 & 71\% \\
\hline
Lua Standard Libraries 5.3 & 87 & 6 & 36 & 129 & 67\% \\
\hline
MD5 & 13 & 0 & 0 & 13 & 100\% \\
\hline
LuaSocket & 109 & 1 & 13 & 123 & 89\% \\
\hline
HTTP Digest & 0 & 0 & 1 & 1 & 0\% \\
\hline
Typical & 1 & 0 & 0 & 1 & 100\% \\
\hline
Modulo 11 & 7 & 0 & 2 & 9 & 78\% \\
\hline
Typed Lua Compiler & 144 & 0 & 10 & 154 & 94\% \\
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for each case study}
\label{tab:evalbycase}
\end{table}

Table \ref{tab:evalbycase} sumarizes our evaluation results for each
case study that we used Typed Lua for typing their members.
We split the members that we typed into three categories:
\emph{easy}, \emph{poly}, and \emph{hard}.
The \emph{easy} category represents the members that we could give
a precise static type for them.
The \emph{poly} category represents the members that we believe that
parametric polymorphism would help give a precise static type for them,
and we had to rely on the dynamic type to type them because our
type system does not support parametric polymorphism.
The \emph{hard} category represents the members that even parametric
polymorphism would not help give a precise static type for them,
and we also had to rely on the dynamic type to type them;
for instance, this category includes functions that require
intersection types to describe their precise static type.
The last column of the table shows the percentage of members that
are under the \emph{easy} category for each case study.
This percentage is our evaluation of Typed Lua, as it
represents how much static typing we could introduce to each one of
our case studies.

Before comparing our type system to related work, we will discuss
each case study in more detail.
For each case study, we will split the evaluation results according
to the modules that each one of them include.
This shall allow us to better discuss the contributions and limitations
of our type system.

\section{Lua Standard Libraries}

The Lua Standard Libraries \citep{luamanual} were our first case study.
We started to think about how we would type them at the same time that
we started to design our type system, as they could give us some hints
on our type system.
And they did: optional parameters and overloading on the return type
are two Lua features that our type system should handle to allow us
typing some of the functions that the standard libraries provide.

All libraries are separated C modules, and we used Typed Lua's description
files to give a statically typed interface to each module.
The \texttt{debug} module is the only one that we did not include in our
evaluation results, because it provides several functions that violate
basic assumptions about Lua code \citep{luamanual}.
For instance, we can use its function \texttt{setlocal} to change the value
of a local variable that is outside of the scope.
Table \ref{tab:evallsl} sumarizes the evaluation results for the Lua Standard Libraries.

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|}
\hline
\textbf{Case study} & \textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
\hline
\multirow{9}{*}{Lua Standard Libraries}
& base & 9 & 1 & 16 & 26 & 35\% \\
\cline{2-7}
& coroutine & 0 & 0 & 6 & 6 & 0\% \\
\cline{2-7}
& package & 5 & 0 & 3 & 8 & 62\% \\
\cline{2-7}
& string & 14 & 0 & 0 & 14 & 100\% \\
\cline{2-7}
& table & 1 & 4 & 1 & 6 & 17\% \\
\cline{2-7}
& math & 28 & 0 & 1 & 29 & 97\% \\
\cline{2-7}
& bit32 & 12 & 0 & 0 & 12 & 100\% \\
\cline{2-7}
& io & 15 & 0 & 6 & 21 & 71\% \\
\cline{2-7}
& os & 10 & 0 & 1 & 11 & 91\% \\
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for Lua Standard Libraries 5.2}
\label{tab:evallsl}
\end{table}

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|}
\hline
\textbf{Case study} & \textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
\hline
\multirow{9}{*}{Lua Standard Libraries}
& base & 9 & 1 & 16 & 26 & 35\% \\
\cline{2-7}
& coroutine & 1 & 0 & 6 & 7 & 14\% \\
\cline{2-7}
& package & 5 & 0 & 3 & 8 & 62\% \\
\cline{2-7}
& string & 14 & 0 & 2 & 16 & 88\% \\
\cline{2-7}
& utf8 & 6 & 0 & 0 & 6 & 100\% \\
\cline{2-7}
& table & 1 & 5 & 1 & 7 & 14\% \\
\cline{2-7}
& math & 26 & 0 & 1 & 27 & 96\% \\
\cline{2-7}
& io & 15 & 0 & 6 & 21 & 71\% \\
\cline{2-7}
& os & 10 & 0 & 1 & 11 & 91\% \\
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for Lua Standard Libraries 5.3}
\label{tab:evallsl53}
\end{table}

@@

The \texttt{base} module was quite hard to type because it
includes several functions that have dynamic behavior.
For instance, the function \texttt{next} traverses all the
fields of a table, but the order that the indices are
enumerated is not specified.
For this reason, it is also difficult to type the function
\texttt{pairs}.
The functions \texttt{getmetatable},  \texttt{rawget}, and
\texttt{rawset} are also examples of functions that their
behavior depend on the C implementation.

@@

We could not type the \texttt{coroutine} module because our
type system does not include the type \emph{thread}.
Lua has one-shot delimited continuations \citep{james2011yield}
in the form of \emph{coroutines} \citep{moura2009rc}, and
effect systems \citep{nielson1999type} are an approach that we
could use to describe control transfers with continuations.
However, for now, coroutines are out of the scope of our type
system, and we use an empty \emph{userdata} declaration
to represent the type \emph{thread}.

@@

We could type all the members of the \texttt{string} module.
Even though we can use the \texttt{string} functions in object-oriented style,
we did not use an \emph{userdata} declaration to handle this feature,
as it would complicate the subtyping rules because the type \texttt{string}
would become a table type instead of a base type.
It was simpler and more starightforward to handle this feature as a
special case in the implementation.
For this reason, we did not include the typing of the \texttt{string} methods
in our evaluation results.

The \texttt{table} module is specially difficult to type because
all of its members require parametric polymorphism.
All the functions of this module either receive or return a list
of elements, and parametric polymorphism would help us to describe
them with a generic type.

@@

The evaluation results show that our type system should include effect
types, parametric polymorphism, and intersection types, as these
features would help to increase the static typing of the Lua Standard Libraries.
Effect types would allow us to type coroutines.
Parametric polymorphism would allow us to define generic function and table types.
Intersection types would allow us to define function types that have
different return types according to their input types.

\section{MD5}

We chose the MD5 library \citep{lmd5} as a case study because it is
simple and contains just one module.
We needed a simple case study to introduce Typed Lua's description
files and \texttt{userdata} declarations.
These Typed Lua's mechanims allow programmers to give statically typed
interfaces to Lua libraries that are written in C.
Table \ref{tab:evalmd5} sumarizes the evaluation results for MD5.

@@

Even tough it was quite straightforward to type the MD5 library,
we found a little difference between its documentation and its static typing.
The documentation suggested that the typing of the function \texttt{update}
should be \texttt{(md5\string_context, string) -> (md5\string_context)},
while a comment in the source code suggested that it should be
\texttt{(md5\string_context, string, string*) -> (md5\string_context)}.
However, while testing it and reading its source code, we noticed that
its actual type is \texttt{(md5\string_context, string*) -> (md5\string_context)},
that is, we can pass zero or more strings to \texttt{update}.

This case study shows that it might be not so obvious to identify which
functions discard extra arguments when we are typing an external library.
In this case study, reading its source code and test script was essential
to confirm the typing of \texttt{update}.

\section{LuaSocket}

LuaSocket \citep{luasocket} is a library that adds network support to Lua,
and it is split into two parts: a core that is written in C and a set of
Lua modules.
The C core provides TCP and UDP support, while the Lua modules provide
support for SMTP, HTTP, and FTP client protocols, MIME encoding,
URL manipulation, and LTN12 filters \citep{nehab2008ltn012}.
We chose LuaSocket as a case study because it is the most popular Lua library.
We used Typed Lua's description files to type both parts.
This is the only case study that we test description files to statically type
a Lua module instead of rewriting it.
Table \ref{tab:evalsocket} sumarizes the evaluation results for LuaSocket.

@@

\section{HTTP Digest}

The HTTP Digest library \citep{luahttpdigest} implements client side
HTTP digest authentication for Lua.
We randonly chose this library as a case study to evaluate Typed Lua
for rewriting an existing Lua module.
Table \ref{tab:evalhttpdigest} summarizes the evaluation results for HTTP Digest.

@@

It is difficul to type the interface of the \texttt{http-digest} module
because it is an extension to the \texttt{http} module from LuaSocket.
The \texttt{http-digest} module only exports the function
\texttt{http-digest.request}, which extends the function
\texttt{http.request} with MD5 authentication.
Like \texttt{http.request}, the return type of \texttt{http-digest.request}
also depends on the type of its parameters.
Its type is either
\texttt{(string) -> (string, number, \{string : string\})?} or
\texttt{(url\string_request) -> (number, number, \{string : string\})?}.

@@

\section{Typical}

Typical \citep{luatypical} is a library that extends the behavior of the
function \texttt{type}. 
We randonly chose this library as a case study to evaluate Typed Lua
for rewriting an existing Lua module.
Table \ref{tab:evaltypical} summarizes the evaluation results for Typical.

@@

\section{Modulo 11}

Modulo 11 \citep{luamod11} is a library that generates and verifies
modulo 11 numbers.
We randonly chose this library as a case study to evaluate Typed Lua
for rewriting an existing Lua module.
Table \ref{tab:evalmod11} summarizes the evaluation results for Typical.

@@

\section{Typed Lua Compiler}

@@

The \texttt{tltype} module implements the types introduced by
Typed Lua and the subtyping and consistent-subtyping relationsr;
it was straightforward to type.

@@

\section{Related Work}

>>>>

We performed some case studies on existing Lua libraries
to evaluate the applicability of our type system.
For each library, we used Typed Lua to either rewrite its modules
or to write statically typed interfaces to its modules through
Typed Lua's description files.
In Section \ref{sec:cases} we present our evaluation results
and discuss some interesting cases.

We also compared our type system to other type systems,
as a way to evaluate our contributions.
In Section \ref{sec:relatedwork} we discuss what are the novel
features introduced by our type system,
what are the features that it shares with other type systems,
and whether we can incorporate some interesting features from
other type systems.

\section{Case Studies}
\label{sec:cases}

We selected seven case studies to evaluate two aspects of Typed Lua:
\begin{enumerate}
\item how precisely it can describe the type of the interface of a module;
\item if it provides guarantees that the code corresponds to the interface.
\end{enumerate}

The Lua Standard Library \citep{luamanual} was our first case study.
We started to think about how we would type its modules at the same time
that we started to design our type system, as it could give us some
hints on our type system.
And it did: optional parameters and overloading on the return type
are two Lua features that our type system should handle to allow us
typing some of the functions that the standard library implements.

The second case study that we chose was the MD5 library \citep{lmd5},
because it is simple and contains just one module.
We needed a simple case study to introduce Typed Lua's description
files and \texttt{userdata} declarations.
These Typed Lua's mechanisms allow programmers to give statically typed
interfaces to Lua libraries that are written in C.

LuaSocket \citep{luasocket} was the third case study that we used to
evaluate Typed Lua.
We chose LuaSocket as a case study because it is the most popular Lua library.

There are three case studies that we randomly selected from the LuaRocks
repository, they are: HTTP Digest \citep{luahttpdigest},
Typical \citep{luatypical}, and Mod 11 \citep{luamod11}.
The first provides client side HTTP digest authentication for Lua.
The second is an extension to the primitive function \texttt{type}.
The third is a generator and checker of modulo 11 numbers.
We randomly selected these libraries to evaluate Typed Lua for
rewriting existing Lua modules.

The Typed Lua compiler is the last case study that we evaluated.
We chose it as a case study because it is a large application.
In addition, bootstrapping the compiler also provides some
confidence on the type system.

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|}
\hline
\textbf{Case study} & \textbf{easy} & \textbf{poly} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
\hline
Lua 5.2 Standard Library & 94 & 5 & 34 & 133 & 71\% \\
\hline
Lua 5.3 Standard Library & 87 & 6 & 36 & 129 & 67\% \\
\hline
MD5 & 13 & 0 & 0 & 13 & 100\% \\
\hline
LuaSocket & 109 & 1 & 13 & 123 & 89\% \\
\hline
HTTP Digest & 0 & 0 & 1 & 1 & 0\% \\
\hline
Typical & 1 & 0 & 0 & 1 & 100\% \\
\hline
Modulo 11 & 7 & 0 & 2 & 9 & 78\% \\
\hline
Typed Lua Compiler & 144 & 0 & 10 & 154 & 94\% \\
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for each case study}
\label{tab:evalbycase}
\end{table}

Table \ref{tab:evalbycase} summarizes our evaluation results for each
of the case studies that we used Typed Lua for typing their members.
We split the members of each case study into three categories:
\emph{easy}, \emph{poly}, and \emph{hard}.
The \emph{easy} category groups the members that we could give
a precise static type.
The \emph{poly} category groups the members that we made minimal
use of the dynamic type as a replacement for the lack of type parameters.
The \emph{hard} category represents the members that either we
had to rely on the dynamic type to type them or we used a static
type that is not precise enough;
for instance, this category includes functions that require
intersection types to describe their precise static type.
The last column of the table shows the percentage of members that
are under the \emph{easy} category for each case study.
This percentage gives a lower bound on how much static type safety
we could add to each one of our case studies.

In the following sections we discuss each case study in more detail.
For each case study, we split the evaluation results according
to the modules that each one of them includes.
This separation by case study allows us to better discuss the
contributions and limitations of our type system.

\subsection{Lua Standard Library}

All of the modules in the standard library are implemented in C, so
we used Typed Lua to type just the interface of each module.
The \texttt{debug} module is the only one that we did not include in our
evaluation results, because it provides several functions that violate
basic assumptions about Lua code \citep{luamanual}.
For instance, we can use its function \texttt{setlocal} to change the value
of a local variable that is not visible in the current scope.
Table \ref{tab:evallsl} summarizes the evaluation results for the Lua Standard Library.

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|}
\hline
\textbf{Case study} & \textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
\hline
\multirow{9}{*}{Lua Standard Library}
& base & 9 & 1 & 16 & 26 & 35\% \\
\cline{2-7}
& coroutine & 0 & 0 & 6 & 6 & 0\% \\
\cline{2-7}
& package & 5 & 0 & 3 & 8 & 62\% \\
\cline{2-7}
& string & 14 & 0 & 0 & 14 & 100\% \\
\cline{2-7}
& table & 1 & 4 & 1 & 6 & 17\% \\
\cline{2-7}
& math & 28 & 0 & 1 & 29 & 97\% \\
\cline{2-7}
& bit32 & 12 & 0 & 0 & 12 & 100\% \\
\cline{2-7}
& io & 15 & 0 & 6 & 21 & 71\% \\
\cline{2-7}
& os & 10 & 0 & 1 & 11 & 91\% \\
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for Lua 5.2 Standard Library}
\label{tab:evallsl}
\end{table}

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|}
\hline
\textbf{Case study} & \textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
\hline
\multirow{9}{*}{Lua Standard Library}
& base & 9 & 1 & 16 & 26 & 35\% \\
\cline{2-7}
& coroutine & 1 & 0 & 6 & 7 & 14\% \\
\cline{2-7}
& package & 5 & 0 & 3 & 8 & 62\% \\
\cline{2-7}
& string & 14 & 0 & 2 & 16 & 88\% \\
\cline{2-7}
& utf8 & 6 & 0 & 0 & 6 & 100\% \\
\cline{2-7}
& table & 1 & 5 & 1 & 7 & 14\% \\
\cline{2-7}
& math & 26 & 0 & 1 & 27 & 96\% \\
\cline{2-7}
& io & 15 & 0 & 6 & 21 & 71\% \\
\cline{2-7}
& os & 10 & 0 & 1 & 11 & 91\% \\
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for Lua 5.3 Standard Library}
\label{tab:evallsl53}
\end{table}

@@

The \texttt{base} module was quite hard to type because it
includes several functions that have a dynamic behavior.
For instance, the function \texttt{next} traverses all the
fields of a table, but the order that the indices are
enumerated is not specified.
For this reason, it is also difficult to type the function
\texttt{pairs}.
The functions \texttt{getmetatable}, \texttt{rawget}, and
\texttt{rawset} are also examples of functions that their
behavior depend on the C implementation.

@@

We could not type the \texttt{coroutine} module because our
type system cannot describe the computational effects of a program.
Lua has one-shot delimited continuations \citep{james2011yield}
in the form of \emph{coroutines} \citep{moura2009rc}, and
effect systems \citep{nielson1999type} are an approach that we
could use to describe control transfers with continuations.
However, for now, coroutines are out of the scope of our type
system, and we use an empty \emph{userdata} declaration
to represent the type \emph{thread}.

@@

We could type all the members of the \texttt{string} module.
Lua programmers often use the functions of the \texttt{string} module
in an object-oriented style, and we treat this feature as a
special case in the implementation.
For this reason, we did not include the typing of the \texttt{string}
methods in our evaluation results.

The \texttt{table} module is specially difficult to type because
almost all of its members require parametric polymorphism.
All the functions of this module either receive or return a list
of elements, and parametric polymorphism would help us to describe
them with a generic type.

@@

The evaluation results show that our type system should include
parametric polymorphism, intersection types, and effect types,
as these features would help us increase the static typing of the
Lua Standard Library.
Parametric polymorphism would allow us to define generic function and table types.
Intersection types would allow us to define overloaded function types.
Effect types would allow us to type coroutines.

\subsection{MD5}

The MD5 library is an OpenSSL based message digest library for Lua.
It contains just the \texttt{md5} module that is written in C,
so we used Typed Lua's description file to type it.
Table \ref{tab:evalmd5} summarizes the evaluation results for MD5.

@@

Even tough it was quite straightforward to type the MD5 library,
we found a little difference between its documentation and its behavior.
The documentation suggested that the typing of the function \texttt{md5.update}
should be \texttt{(md5\string_context, string) -> (md5\string_context)},
while a comment in the source code suggested that it should be
\texttt{(md5\string_context, string, string*) -> (md5\string_context)}.
However, while testing it and reading its source code, we noticed that
its actual type is \texttt{(md5\string_context, string*) -> (md5\string_context)},
that is, we can pass zero or more strings to \texttt{md5.update}.

This case study shows that it might be not so obvious to identify which
functions discard extra arguments when we are typing an external library.
In this case study, reading its source code and test script was essential
to confirm the typing of \texttt{md5.update}.

\subsection{LuaSocket}

LuaSocket is a library that adds network support to Lua,
and it is split into two parts: a core that is written in C and a set of
Lua modules.
The C core provides TCP and UDP support, while the Lua modules provide
support for SMTP, HTTP, and FTP client protocols, MIME encoding,
URL manipulation, and LTN12 filters \citep{nehab2008ltn012}.
We used Typed Lua's description files to type both parts.
This is the only case study that we test description files to statically type
a Lua module instead of rewriting it.
Table \ref{tab:evalsocket} summarizes the evaluation results for LuaSocket.

@@

\subsection{HTTP Digest}

The HTTP Digest library implements client side HTTP digest authentication for Lua.
Table \ref{tab:evalhttpdigest} summarizes the evaluation results for HTTP Digest.

@@

It is difficult to type the interface of the \texttt{http-digest} module
because it is an extension to the \texttt{http} module from LuaSocket.
The \texttt{http-digest} module only exports the function
\texttt{http-digest.request}, which extends the function
\texttt{http.request} with MD5 authentication.
Like \texttt{http.request}, the return type of \texttt{http-digest.request}
also depends on the type of its parameters.
Its type is either
\texttt{(string) -> (string, number, \{string : string\})?} or
\texttt{(url\string_request) -> (number, number, \{string : string\})?}.

@@

\subsection{Typical}

Typical is a library that extends the behavior of the function \texttt{type}. 
Table \ref{tab:evaltypical} summarizes the evaluation results for Typical.

@@

\subsection{Modulo 11}

Modulo 11 is a library that generates and verifies modulo 11 numbers.
Table \ref{tab:evalmod11} summarizes the evaluation results for Typical.

@@

\subsection{Typed Lua Compiler}

@@

The \texttt{tltype} module implements the types introduced by
Typed Lua and the subtyping and consistent-subtyping relations;
it was straightforward to type.

@@

\section{Related Work}
\label{sec:relatedwork}

### 2014-11-07 doc/thesis/introduction.tex
Splitting the evaluation chapter into two sections: case studies and related work

Lua is a small scripting language, so designing an optional type
system for Lua may shed some light on optional type systems for
scripting languages in general.
Lua provides mechanisms for organizing a program in modules and
writing object-oriented programs, but it does not set policy on how
these features should behave, due to its use as an embedded and
extension language.
Thus, implementing an optional type system for Lua offers Lua
programmers one way to obtain most of the advantages of static typing
without compromising the simplicity and flexibility of dynamic typing.

So far, Typed Lua is a strict superset of Lua that
provides optional type annotations, compile-time type checking, and
class-based object-oriented programming through the definition of
classes, interfaces, and modules.
In practice, we implement Typed Lua as a programming language that
extends Lua syntax to add optional type annotations and standard
constructions for structuring Lua code.
The compiler uses static types to perform compile-time type
checking, but also allows Lua code to coexist with Typed Lua code,
and generates Lua code that runs in unmodified Lua implementations.
We have an implementation of the Typed Lua compiler that is
available online\footnote{https://github.com/andremm/typedlua}.

Typed Lua intended use is as an application language, and we view
that policies for organizing a program in modules and writing
object-oriented programs should be part of the language and
enforced by its optional type system.
An application language is a programming language that helps
programmers develop applications from scratch until these
applications evolve to complex systems rather than just scripts.

@@

In Chapter \ref{chap:review} we review the literature about static and
dynamic typing in the same language, which includes the discussion
about the particularities between optional type systems and gradual
typing.
In Chapter \ref{chap:typedlua} we use code examples to present the
design of Typed Lua.
In Chapther \ref{chap:system} we use typing rules to present the
formalization of Typed Lua.
In Chapter \ref{chap:evaluation} we discuss some case studies that
we used to evaluate our design.
In Chapter \ref{chap:conc} we outline our contributions.

>>>>

Lua is a small imperative language with first-class functions
(with proper lexical scoping) where the only data structure
mechanism is the \emph{table}, an associative array that can
represent arrays, records, maps, modules, objects, etc.
Tables also have syntactic sugar and metaprogramming support
through operator overloading built into the language.
Unlike other scripting languages, Lua has very limited coercion
among different data types.
We believe that designing an optional type system for Lua may
shed some light on optional type systems for scripting languages
in general, as Lua is a small scripting language that shares
some features with other scripting languages such as JavaScript.

Lua prefers to provide mechanisms instead of fixed policies due
to its primary use that has always been as an embedded language
for configuration and extension of other applications.
This means that even features such as a module system and
object orientation are a matter of convention instead of built
into the language.
The result is a fragmented ecosystem of libraries, and different
ideas among Lua programmers on how they should use the language
features, and how they should structure programs.

The lack of standard policies is a challenge for the design of
an optional type system for Lua.
For this reason, we are not relying just on the semantics of
the language to design our type system.
We also run a mostly automated survey of Lua idioms used in a
large corpus of Lua libraries, and it also helped us on the
design of Typed Lua.

So far, Typed Lua is a Lua extension that allows statically typed
code to coexist and interact with dynamically typed code
through optional type annotations, and it also adds standard
constructions for structuring Lua code.
The Typed Lua compiler warns programmers about type errors,
but always generates Lua code that runs in unmodified Lua implementations.
Programmers can enjoy some of the benefits of static types
even without converting existing Lua modules to Typed Lua:
they can export a statically typed interface to a dynamically typed module,
and statically typed users of the module can use the Typed Lua compiler
to check their use of the module.
Thus, implementing an optional type system for Lua offers Lua
programmers one way to obtain most of the advantages of static typing
without compromising the simplicity and flexibility of dynamic typing.
We have an implementation of the Typed Lua compiler that is
available online\footnote{https://github.com/andremm/typedlua}.

Typed Lua intended use is as an application language, and we view
that policies for organizing a program in modules and writing
object-oriented programs should be part of the language and
enforced by its optional type system.
An application language is a programming language that helps
programmers develop applications from scratch until these
applications evolve to complex systems rather than just scripts.
We shall see that Typed Lua introduces the refinement of
tables to support the common way that Lua programmers use
them to represent both modules and objects.

@@

In Chapter \ref{chap:review} we review the literature about static
and dynamic typing in the same language, we discuss the differences
between optional type systems and gradual typing, and we also
present the results of our survey on Lua idioms.
In Chapter \ref{chap:typedlua} we use code examples to present the
design of Typed Lua.
In Chapter \ref{chap:system} we use typing rules to present the
formalization of Typed Lua.
In Chapter \ref{chap:evaluation} we discuss some case studies that
we used to evaluate our design, and we also compare our design
to related work.
In Chapter \ref{chap:conc} we outline our contributions.
